Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-06-09 paper.tex
last edits #7 summarizing discussion +diagnostic plots

Using the imputations from \eqref{eq:estY10}, we estimate
$\EE\delta(1)$ as
\begin{equation*}
\widehat{\EE\delta(1)}=\widehat{\EE Y(Z=1)}-\widehat{\EE
  Y(1,0)}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,0)
\end{equation*}
where $n_T=\sum_i Z_i$, the number of subjects in the treatment
group.
Since $Y_i=\hat{Y}_i(1,0)$ whenever $H_i=0$, the summand is non-zero
only when $H_i=1$.
Further, note that when multiple $H=1$ subjects share a
matched set, they also share imputations $\hat{Y}(1,0)$.
Re-writing $\widehat{\EE\delta(1)}$ in terms of matched sets $M=m$,
\begin{align}
\widehat{\EE\delta(1)}&=\frac{1}{n_T}\displaystyle\sum_m\displaystyle\sum_{i:H_i=1,M_i=m}
                        Y_i-\bar{Y}_{M=m_i,H=0} \nonumber \\
&=\frac{1}{n_T}\displaystyle\sum_m
  n_{1m}(\bar{Y}_{M=m_i,H=1}-\bar{Y}_{m=m_i,H=0}) \nonumber \\
&=\frac{1}{n_T}\displaystyle\sum_mn_{1m}\hat{\tau}_m=\frac{\sum_iH_i}{n_T}\widehat{TOT} \label{eq:indTot1}
\end{align}
where $n_{1m}$ and $\hat{\tau}_m$ were defined in Section
\ref{sec:obsEst} as the number of $H=1$ subjects in match $m$ and the
difference in the sample mean of $Y$ between $H=1$ and $H=0$ subjects
in match $m$, respectively, and $\hat{TOT}$ is the TOT estimate,
\eqref{eq:tauWeightedMean} with $w_m=n_{1m}$.
The upshot of \eqref{eq:indTot1} is that under matched ignorability,
$\EE\delta(0)$ can be estimated as the TOT times the proportion of the
treatment group with $H=1$.

@@

In the hint study, the definition of $H$ implies that
$Pr(H=1)=\sum_iH_i/n_T=$0.3.
Therefore, our estimate of the average indirect effect is
$\widehat{\EE \delta(1)}=$0.3$\widehat{TOT}$.
In Table \ref{tab:matchResults}, we estimated the TOT as
-0.14 with a
standard error of 0.06.
Thefore, we estimate an average indirect effect of -0.04, with
a standard error of 0.02.

It is worth repeating that these estimates assume matched ignorabilty---no confounding between
$H$ and $Y$ within matched sets.
With this important caveat, this analysis suggests that the
wide availability of a hints
actually \emph{lowers} the treatment effect, that CTA1 works not
because of hints, but despite them.

@@

What would the effect be were all students to request few hints?
To answer that question, we estimate average direct effects,
$\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]=\EE[Y(1,0)-Y(0,0)]$.\footnote{This
  is equivalent to the ``controlled direct effect,'' $CDE(0)$
  e.g. \citet[][p. 57]{vanderweele2015explanation}.}
In principal, this could be estimated as the difference in sample
means of $\hat{Y}(1,0)$ in the treatment group and observed $Y=Y(0,0)$
in the control group.
In general, if we let
\begin{equation*}
\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)Y(0,0)
\end{equation*}
then we can estimate $\EE[\xi(0)]$ as the ATE of treatment assignment
$Z$ on $\tilde{Y}$.
Just like estimates of the ATE of $Z$ on $Y$, this estimate should
account for the experimental design: a paired, group-randomized trial
with students clustered in schools and schools randomized within
pairs.

For the sake of simplicity, we will use an OLS regression estimator,
with cluster-robust standard errors \citep{pustejovsky2018small},
clustered at the school level.
(Since in Section \ref{sec:observational} subjects were matched
within schools, school-level cluster-robust standard errors also account for the
dependence of $\hat{Y}(1,0)$ between matched treated students.)
Specifically:
\begin{equation}\label{eq:directEffReg}
\tilde{Y}_i=\alpha_{p[i]}+\hat{\EE\xi}Z_i+\epsilon_i
\end{equation}
where $\alpha_{p[i]}$ is a fixed intercept for the randomization pair
of subject $i$'s school, $\hat{\EE\xi}$ is an estimate of $\EE\xi$ and $\epsilon_i$ is a regression error.
Using the \texttt{clubSandwich} package in \texttt{R}
\citep{clubsandwich} to estimate standard errors, and 95\% confidence intervals, we estimated a direct
effect of
0.14$\pm$0.25.
Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
0.18$\pm$0.21.

Table \ref{tab:mediation} summarizes our mediation results: assuming matched ignorability, assignment to CTA1 increases student test scores despite the availablity of hints (an indirect effect of -0.04 standard deviations), and the effect due to CTA1's other mechanisms is 0.18.

@@

In an RCT, identification of these varying effects relies on
randomization of treatment assignment,
rather than confounding control or assumptions about
unmeasured confounders.
Unlike the observational study or mediation analysis, PS in an RCT requires no assumptions about unmeasured confounding.
PS analysis requires data on outcomes
(e.g. posttest scores) in both tretment arms, as well as data on
implementation.
In classical PS
\citep[e.g.][]{page2012principal,feller2016compared}, the intermediate
variable, implementation, is discrete or categorical, leading to
discrete strata.
Other approachs \citep[e.g.]{gilbertHudgens,jin2008principal} allow
the intermediate variable to be continuous.
The method we use here is based on \citet{aoas}, in which the variable
defining principal strata is latent---specifically, the ``ability''
parameter is \eqref{eq:rasch}.
For this approach, implementation is measured with a series of binary
measurements for each student, with each measurement taken from a
specific problem or section.
Other implementation data structures may call for different
measurement models.
Finally, covariates may not be strictly necessary for PS, but they can be extremely helpful.

Below, we describe a Bayesian PS model fit in one step.
However, the full process of PS analysis is a much more involved
procedure.
After a formal introduction in Section \ref{sec:psIntro},
\S~\ref{sec:psModel} describes a set of sub-models for student hint
requests, posttest scores, and treatment effects.
Ideally, each of these models should be carefully tailored to each
data application, and each must be rigorously tested.
Model checking includes examination of residual data plots, fake data
simulation, and fitting a range of alternative specifications in a
sensitivity analysis.
More details for checking PS models can be found in \citet{aoas},
\citet{psTutorial}, and for Bayesian models in general in
\citet{gelman2014bayesian} and \citet{mcelreath2020statistical}.

@@

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}(r)$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
Eighty-eight percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of 0.88 that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
The right hand panel plots one posterior draw of $\eta(1)$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.

>>>>

Using the imputations from \eqref{eq:estY10}, we estimate
$\EE\delta(1)$ as
\begin{equation*}
\widehat{\EE\delta(1)}=\widehat{\EE Y(Z=1)}-\widehat{\EE
  Y(1,0)}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,0)
\end{equation*}
where $n_T=\sum_i Z_i$, the number of subjects in the treatment
group.
Since $Y_i=\hat{Y}_i(1,0)$ whenever $H_i=0$, the summand is non-zero
only when $H_i=1$.
Further, note that when multiple $H=1$ subjects share a
matched set, they also share imputations $\hat{Y}(1,0)$.
Re-writing $\widehat{\EE\delta(1)}$ in terms of matched sets $M=m$,
\begin{align}
\widehat{\EE\delta(1)}&=\frac{1}{n_T}\displaystyle\sum_m\displaystyle\sum_{i:H_i=1,M_i=m}
                        Y_i-\bar{Y}_{M=m_i,H=0} \nonumber \\
&=\frac{1}{n_T}\displaystyle\sum_m
  n_{1m}(\bar{Y}_{M=m_i,H=1}-\bar{Y}_{m=m_i,H=0}) \nonumber \\
&=\frac{1}{n_T}\displaystyle\sum_mn_{1m}\hat{\tau}_m=\frac{\sum_iH_i}{n_T}\widehat{TOT} \label{eq:indTot1}
\end{align}
where $n_{1m}$ and $\hat{\tau}_m$ were defined in Section
\ref{sec:obsEst} as the number of $H=1$ subjects in match $m$ and the
difference in the sample mean of $Y$ between $H=1$ and $H=0$ subjects
in match $m$, respectively, and $\widehat{TOT}$ is the TOT estimate,
\eqref{eq:tauWeightedMean} with $w_m=n_{1m}$.
The upshot of \eqref{eq:indTot1} is that under matched ignorability,
$\EE\delta(0)$ can be estimated as the TOT times the proportion of the
treatment group with $H=1$.

@@

In the hint study, the definition of $H$ implies that
$Pr(H=1)=\sum_iH_i/n_T=$0.3.
Therefore, our estimate of the average indirect effect is
$\widehat{\EE \delta(1)}=$0.3$\widehat{TOT}$.
In Table \ref{tab:matchResults}, we estimated the TOT as
-0.14 with a
standard error of 0.06.
That implies an average indirect effect of -0.04, with
a standard error of 0.02.

It is worth repeating that these estimates assume matched ignorability---no confounding between
$H$ and $Y$ within matched sets.
With this important caveat, this analysis suggests that the
wide availability of a hints
actually \emph{lowers} the treatment effect, that CTA1 works not
because of hints, but despite them.

@@

What would the effect be were all students to request few hints?
To answer that question, we estimate average direct effects,
$\EE[\xi(0)]=\EE[Y(1,H(0))-Y(0,H(0))]=\EE[Y(1,0)-Y(0,0)]$.\footnote{This
  is equivalent to the ``controlled direct effect,'' $CDE(0)$
  e.g. \citet[][p. 57]{vanderweele2015explanation}.}
In principal, this could be estimated as the difference in sample
means of $\hat{Y}(1,0)$ in the treatment group and observed $Y=Y(0,0)$
in the control group.
In general, if we let
\begin{equation*}
\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)Y(0,0)
\end{equation*}
then we can estimate $\EE[\xi(0)]$ as the ATE of treatment assignment
$Z$ on $\tilde{Y}$.
Just like estimates of the ATE of $Z$ on $Y$, this estimate should
account for the experimental design: a paired, group-randomized trial
with students clustered in schools and schools randomized within
pairs.

For the sake of simplicity, we will use an OLS regression estimator,
with cluster-robust standard errors \citep{pustejovsky2018small},
clustered at the school level.
(Since in Section \ref{sec:observational} subjects were matched
within schools, school-level cluster-robust standard errors also account for the
dependence of $\hat{Y}(1,0)$ between matched treated students.)
Specifically, we regress:
\begin{equation}\label{eq:directEffReg}
\tilde{Y}_i=\alpha_{p[i]}+\widehat{\EE[\xi(0)]}Z_i+\epsilon_i
\end{equation}
where $\alpha_{p[i]}$ is a fixed intercept for the randomization pair
of subject $i$'s school and $\epsilon_i$ is a regression error.
Using the \texttt{clubSandwich} package in \texttt{R}
\citep{clubsandwich} to estimate standard errors, and 95\% confidence intervals, we estimated a direct
effect of
0.14$\pm$0.25.
Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
0.18$\pm$0.21.

Table \ref{tab:mediation} summarizes our mediation results: assuming matched ignorability, assignment to CTA1 increases student test scores despite the availability of hints (an indirect effect of -0.04 standard deviations), and the effect due to CTA1's other mechanisms is 0.18.

@@

In an RCT, identification of these varying effects relies on
randomization of treatment assignment,
rather than confounding control or assumptions about
unmeasured confounders.
Unlike the observational study or mediation analysis, PS in an RCT requires no assumptions about unmeasured confounding.
PS analysis requires data on outcomes
(e.g. posttest scores) in both treatment arms, as well as data on
implementation.
In classical PS
\citep[e.g.][]{page2012principal,feller2016compared}, the intermediate
variable, implementation, is discrete or categorical, leading to
discrete strata.
Other approaches \citep[e.g.]{gilbertHudgens,jin2008principal} allow
the intermediate variable to be continuous.
The method we use here is based on \citet{aoas}, in which the variable
defining principal strata is latent---specifically, the ``ability''
parameter is \eqref{eq:rasch}.
For this approach, implementation is measured with a series of binary
measurements for each student, with each measurement taken from a
specific problem or section.
Other implementation data structures may call for different
measurement models.
Finally, covariates may not be strictly necessary for PS, but they can be extremely helpful.

Below, we describe a Bayesian PS model fit in one step.
However, the full process of PS analysis is a much more involved
procedure.
After a formal introduction in Subsection \ref{sec:psIntro},
Subsection \ref{sec:psModel} describes a set of sub-models for student hint
requests, posttest scores, and treatment effects.
Ideally, each of these models should be carefully tailored to each
data application, and each must be rigorously tested.
Model checking includes examination of residual data plots, fake data
simulation, and fitting a range of alternative specifications in a
sensitivity analysis.
More details for checking PS models can be found in \citet{aoas},
\citet{psTutorial}, and for Bayesian models in general in
\citet{gelman2014bayesian} and \citet{mcelreath2020statistical}.

@@

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}(r)$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
Eighty-eight percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of 0.75
that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
The right hand panel plots one posterior draw of $\eta(1)$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.
